Parallel Hierarchical Clustering using \\ Rank-Two Nonnegative Matrix Factorization \thanks{Identify applicable funding agency here. If none, delete this.}

### 2020-07-04 latex/paper.tex
added description of rank-2 nnls algorithm

\usepackage{cite,doi}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithm,algpseudocode}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{cleveref}
\usepackage{tikz}
\usepackage{ourmacros}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

@@

\begin{itemize}
	\item NMF, BCD, rank-2 NMF
	\item PLANC
\end{itemize}

@@

\begin{algorithm}
\caption{Rank-2 Nonnegative Least Squares Solve \cite{KP13}}
\label{alg:r2nmf}
\begin{algorithmic}[1]
	\Require{$\M{C}$ is $n\times 2$ and $\M{G}$ is $2\times 2$ and s.p.d.}
	\Function{$\M{H} =$ Rank2-NLS-Solve}{$\M{C},\M{G}$}
		\State $\M{H} = \M{C}\M{G}^{-1}$ \hfill \Comment{Solve unconstrained system}
		\For{$i=1$ to $n$}
			\If{$\ME{H}{i1} < 0$ or $\ME{H}{i2} < 0$}
				\State \Comment{Choose between single-variable solutions}
				\If{$\ME{H}{i1} \sqrt{\ME{G}{11}}  < \ME{H}{i2} \sqrt{\ME{G}{22}} $}
					\State $\ME{H}{i1} = 0$
					\State $\ME{H}{i2} = \ME{C}{i2} / \ME{G}{22}$
				\Else
					\State $\ME{H}{i1} = \ME{C}{i1} / \ME{G}{11}$
					\State $\ME{H}{i2} = 0$
				\EndIf
			\EndIf
		\EndFor
	\EndFunction
	\Ensure{$\displaystyle \M{H} = \argmin_{\M[\bar]{H} \geq \M{0}} \|\M{A}-\M{W}\M[\bar]{H}^\Tra\|$ is $n\times 2$ with $\M{C}=\M{A}^\Tra \M{W}$ and $\M{G}=\M{W}^\Tra\M{W}$}
\end{algorithmic}
\end{algorithm}

>>>>

\usepackage{cite,doi}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithm,algpseudocode}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{cleveref}
\usepackage{tikz}
\usepackage{mathtools}
\usepackage{ourmacros}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

@@

\begin{itemize}
	\item NMF, BCD, NNLS
	\item PLANC
\end{itemize}

@@

Using the 2-block BCD approach for a rank-2 NMF yields NNLS subproblems of the form $\displaystyle \min_{\M[\bar]{H} \geq \M{0}} \|\M{W}\M[\bar]{H}^\Tra-\M{A}\|$ and $\min_{\M[\bar]{W} \geq \M{0}} \|\M{H}\M[\bar]{W}^\Tra-\M{A}^\Tra\|$.
In each case, the columns of the transposed variable matrix can be computed independently.
Considering the $i$th row of $\M[\bar]{H}$, for example, the NNLS problem to solve is
\begin{align*}
\begin{multlined}
\min_{\ME[\bar]{H}{i,1},\ME[\bar]{H}{i,2} \geq 0} \left\|\begin{bmatrix} \V{w}_1 & \V{w}_2 \end{bmatrix} \begin{bmatrix}\ME[\bar]{H}{i,1} \\ \ME[\bar]{H}{i,2} \end{bmatrix} - \V{A}_i \right\| \\ = \min_{\ME[\bar]{H}{i,1},\ME[\bar]{H}{i,2} \geq 0} \left\|\ME[\bar]{H}{i,1} \V{w}_1 + \ME[\bar]{H}{i,2} \V{w}_2 - \V{a}_i \right\|
\end{multlined}
\end{align*}
where $\V{w}_1$ and $\V{W}_2$ are the two columns of $\M{W}$ and $\V{a}_i$ is the $i$ column of $\M{A}$.
We note that there are four possibilities of solutions, as each of the two scalar variables may be positive or zero.

As shown by Kuang and Park \cite{KP13}, determining which of the four possible solutions is feasible and optimal can be done efficiently by exploiting the following properties:
\begin{itemize}
	\item if the solution to the unconstrained least squares problem admits two positive values, it is the optimal solution to the nonnegatively constrained problem,
	\item if $\M{W}$ and $\M{A}$ are both nonnegative, then the candidate solution with two zero values is never (uniquely) optimal and can be discarded, and
	\item if the solution to the unconstrained problem does not admit two positive values, the better of the two remaining solutions can be determined by comparing $\V{a}_{j}^\Tra\V{w}_1 / \|\V{w}_1\|$ and $\V{a}_{j}^\Tra\V{w}_2 / \|\V{w}_2\|$.
\end{itemize}
If the unconstrained problem is solved via the normal equations, then the temporary matrices computed for the normal equations ($\M{W}^\Tra\M{W}$ and $\M{A}^\Tra\M{W}$) can be re-used to determine the better of the two single-positive-variable solutions.

\Cref{alg:r2nmf} implements this strategy for all rows of $\M{H}$ simultaneously.
It takes as input the matrices $\M{C}=\M{A}^\Tra\M{W}$ and $\M{G}=\M{W}^\Tra\M{W}$, first solves the normal equations for the unconstrained problem, and then determines between the two alternate possibilities as necessary.
\GB{To avoid two passes through the data, we could combine the solve and the update if necessary.  It might also help to pre-compute the inverse of $\M{G}$...  Also Lawton, please check the details of the pseudocode!}

\begin{algorithm}
\caption{Rank-2 Nonnegative Least Squares Solve \cite{KP13}}
\label{alg:r2nmf}
\begin{algorithmic}[1]
	\Require{$\M{C}$ is $n\times 2$ and $\M{G}$ is $2\times 2$ and s.p.d.}
	\Function{$\M{H} =$ Rank2-NLS-Solve}{$\M{C},\M{G}$}
		\State $\M{H} = \M{C}\M{G}^{-1}$ \hfill \Comment{Solve unconstrained system}
		\For{$i=1$ to $n$}
			\If{$\ME{H}{i1} < 0$ or $\ME{H}{i2} < 0$}
				\State \Comment{Choose between single-variable solutions}
				\If{$\ME{C}{i1} / \sqrt{\ME{G}{11}}  < \ME{C}{i2} / \sqrt{\ME{G}{22}} $}
					\State $\ME{H}{i1} = 0$
					\State $\ME{H}{i2} = \ME{C}{i2} / \ME{G}{22}$
				\Else
					\State $\ME{H}{i1} = \ME{C}{i1} / \ME{G}{11}$
					\State $\ME{H}{i2} = 0$
				\EndIf
			\EndIf
		\EndFor
	\EndFunction
	\Ensure{$\displaystyle \M{H} = \argmin_{\M[\bar]{H} \geq \M{0}} \|\M{A}-\M{W}\M[\bar]{H}^\Tra\|$ is $n\times 2$ with $\M{C}=\M{A}^\Tra \M{W}$ and $\M{G}=\M{W}^\Tra\M{W}$}
\end{algorithmic}
\end{algorithm}
